---
title: "County Cluster R"
output: html_notebook
---

This is a rewrite of my County Cluster notebook - originally written in Python in Jupyter Notebook - using R. The code may not be pretty, buy my goal is simply to practice R and demonstrate basic competence. 

Loading the tidyverse
```{r}
library(tidyverse)
library(maps)
library(mapproj)
```

Read in datasets
```{r}
ed_data <- read_csv("../data/Education.csv", locale = locale(encoding = "Latin1"))
pop_data <- read_csv("../data/PopulationEstimates.csv", locale = locale(encoding = "Latin1"))
pov_data <- read_csv("../data/PovertyEstimates.csv", locale = locale(encoding = "Latin1"))
unm_data <- read_csv("../data/Unemployment.csv", locale = locale(encoding = "Latin1"))
```
Taking a look at the education data
```{r}
head(ed_data)
```
And the ed_data dimensions
```{r}
dim(ed_data)
```
Now looking at population data - it had a problem when reading in, we'll have to interrogate that more later
```{r}
head(pop_data)
```
pop_data dimensions
```{r}
dim(pop_data)
```
And now poverty data
```{r}
head(pov_data)
```
poverty dimensions
```{r}
dim(pov_data)
```
And finally, unemployment data
```{r}
head(unm_data)
```
and unemployment data dimensions
```{r}
dim(unm_data)
```
A quick look at the summaries of each
```{r}
#summary(ed_data)
```
```{r}
#summary(pop_data)
```

```{r}
#summary(pov_data)
```


```{r}
#summary(unm_data)
```
So it turns out that while summary() in R returns much the same information as .describe() does in pandas, it does it in a really ungainly format. I don't think I'll be doing that again. I've commented out the summaries to avoid clogging the notebook. 
At this point in pandas I listed out the data types in each dataframe, but here those have already been displayed on the head(). I also loaded in a shape file with the geometries of the counties, but this was ultimately only used for plotting and the maps package in R already covers US county mapping, so I won't be using the shape file for this. 
```{r}

```
